Formalizing Stack Safety as a Security Property

### 2022-08-21 submission5/onlyformalism.tex
Working on pseudoinstruction formalism

We begin with a RISC-V-like machine enhanced with PIPE, a tag-based reference
monitory. We extend this machine with a {\it security overlay semantics}:
a semantics in which calls, returns, and other operations of interest are condensed
into atomic {\it macro-instructions}, which update an additional security context.
Our properties are defined in terms of the security context at the end of an appropriate
macro-instruction.

@@

Collectively addresses and registers are {\em state elements} \(\component\)
in the set \(\COMPONENTS ::= \WORDS + \REGS\).
%
A {\em machine state} \(\mach \in \COMPONENTS \rightarrow \WORDS \)
is a mapping from state elements to values. For convenience, we abbreviate
multiple updates \(m[\component_0 \mapsto w][\component_1 \mapsto w]\dots\)
as \(m \llbracket \component \mapsto w | P ~ \component\rrbracket\), where
\(P\) is some predicate that defines the set of elements to update.

@@

Execution is modeled by a partial step function between machine states,
written \(\mach, \context \stepsto \mach', \context'\).
If \(\mach[\PCname]\) decodes to an instruction, then
\(\mach, \context \stepsto \mach', \context\) if \(\mach\) steps to \(\mach'\)
under the RISC-V semantics.
If, for some \(j\), \(\mach[\PCname + i]\) decodes to \(j_i\) for all \(i\) between 0 and
\(\mathit{size}(j)\), then the step is via a special multi-step. We present these steps
as separate step relations: \(\cdot \harpoonunder{j}_M \cdot \in \MACHS \times J \times \MACHS\)
to step the machine state,
and \(\cdot, \cdot \stepstounder{j}_C \cdot \in \MACHS \times \CONTEXTS \times J \times \CONTEXTS\)
for the security context. We then combine them:

\judgmenttwo[]
            {\(m \harpoonunder{j}_M m'\)}
            {\(m',c \stepstounder{j}_C c'\)}
            {\(m,c \stepsto m',c'\)}

@@

Then, on enter, the callee stores the stack pointer and allocates new stack space.
[TODO: also have it do the saving of other callee-save, little complicated.]

\vspace{\abovedisplayskip}
\judgmentbrbr[EnterM]
           {\(j = \mathit{decode}^* ~ m ~ m[pc] = \mathit{enter} ~ tr\)}
           {\(m' = m[m[sp] \mapsto m[rra]]\)}
           {\(m'' = m'[sp \mapsto m[sp]+n+1]]\)}
           {\(m \harpoonunder{j}_M m''[pc \mapsto m[pc]+4*\mathit{size}(j)]\)}
\vspace{\belowdisplayskip}

Finally, on return, the program counter is restored from the stack, and the callee's
frame is deallocated.

\vspace{\abovedisplayskip}
\judgmentbr[ReturnM]
             {\(j = \mathit{decode}^* ~ m ~ m[pc] = \mathit{return} ~ n\)}
             {\(m' = m[sp \mapsto m[sp]-n-1]\)}
             {\(m \harpoonunder{j}_M m'[pc \mapsto m'[sp]]\)}
\vspace{\belowdisplayskip}

@@

[TODO]

\paragraph*{Machine Traces}
\label{sec:traces}

A {\em machine trace} is a nonempty, finite or infinite sequence
of states, ranged over by \(\MPCT\) and \(\NPCT\).
We use ``\(\notfinished{}{}\)'' to represent ``cons'' for traces, reserving ``::''
for list-cons.

The ``trace-of'' operator, written \(\mpcstatename \hookrightarrow \MPCT\),
coinductively relates an initial state with the trace of states
produced by repeated application of \(\stepstocon\):
%
\vspace{-\bigskipamount}
\begin{center}
\begin{minipage}[t]{.6\columnwidth}
\judgmenttwo{\(\mpcstatename \stepstocon \mpcstatename'\)}
            {\(\mpcstatename' \hookrightarrow \MPCT\)}
            {\(\mpcstatename \hookrightarrow \notfinished{\mpcstatename'}{\MPCT}\)}%
\end{minipage}
\begin{minipage}[t]{.35\columnwidth}
\judgment%[Default]
         {\(\not\exists \mpcstatename'. \mpcstatename \stepstocon \mpcstatename'\)}
         {\(\mpcstatename \hookrightarrow \mpcstatename\)}
\end{minipage}
\end{center}
\vspace{-\smallskipamount}

@@

\paragraph*{Until}
The operation \(\PUT ~ f ~ \MPCT\) takes a trace \(\MPCT\)
and a predicate on states \(f \subseteq \MPCS\) and gives the prefix of
\(\MPCT\) ending with the first trace entry on which \(f\) holds.

\vspace{-\medskipamount}
\begin{center}
  \begin{minipage}[b]{.5\columnwidth}
    \judgment{\(f ~ \mpcstatename\)}
             {\(\PUT ~ f ~ (\notfinished{\mpcstatename}{\MPCT}) = \mpcstatename\)}
  \end{minipage}
%
  \begin{minipage}[b]{.3\columnwidth}
    \judgment{}
             {\(\PUT ~ f ~ \mpcstatename = \mpcstatename\)}
%
  \end{minipage}
\end{center}
    \judgmenttwo{\(\neg f ~ \mpcstatename\)}{\(\PUT ~ f ~ \MPCT = \MPCT'\)}
                {\(\PUT ~ f ~ (\notfinished{\mpcstatename}{\MPCT}) = \notfinished{\mpcstatename}{\MPCT'}\)}

If the resulting trace is a strict prefix of the parameter, then the
predicate must hold on its final entry. If the predicate never holds, then
\(\PUT\) is the identity function.

We will frequently take a prefix of the trace from an initial state up
through the first state where some condition holds on the machine state.
This can easily be implemented using \(\PUT\), and we provide special notation
for convenience.
%
If \(f\) is a predicate on machine states, we define \(\mpcstatename
\hookrightarrow \MPCT | f\) (pronounced ``\(\MPCT\) is the prefix of
running \(\mpcstatename\) up to \(f\)'') thus:
%
\begin{center}
\judgmenttwo{\(\mpcstatename \hookrightarrow \MPCT'\)}
            {\(\PUT ~ f ~ \MPCT' = \MPCT\)}
            {\(\mpcstatename \hookrightarrow \MPCT | f\)}
\end{center}

\paragraph*{Context Segments}

In particular, we wish to formulate properties around consecutive subtraces
in which the context obeys a given predicate. This will commonly be used to extract
individual calls from a trace by taking subtraces in which the stack is
at or above a given depth. If \(f\) is again a predicate on states, then
we define \(\mpcstatename \hookrightarrow \MPCT \langle f \rangle\) (pronounced
``\(\mpcstatename\) segmented by \(f\) yields a sub-trace \(\MPCT\)'') thus:
\ifspace{\bcp{Throughout, we could save space and improve beauty by removing some of the
vertical space between inference rules.}\fi

\begin{center}
  \judgmenttwo[]
              {\(f ~ \mpcstatename\)}
              {\(\mpcstatename \hookrightarrow \MPCT | \neg f\)}
              {\(\mpcstatename \hookrightarrow \MPCT \langle f \rangle\)}
  \judgmentthree[]
                {\(f ~ \mpcstatename\)}
                {\(\mpcstatename \hookrightarrow \MPCT | \neg f\)}
                {\(\last(\MPCT) \hookrightarrow \MPCT' \langle f \rangle\)}
                {\(\mpcstatename \hookrightarrow \MPCT' \langle f \rangle\)}
  \judgmentthree[]
                {\(\neg f ~ \mpcstatename\)}
                {\(\mpcstatename \hookrightarrow \MPCT | f\)}
                {\(\last(\MPCT) \hookrightarrow \MPCT' \langle f \rangle\)}
                {\(\mpcstatename \hookrightarrow \MPCT' \langle f \rangle\)}
\end{center}

@@

We often want to quantify over all traces from any initial state so that our
properties apply to the system as a whole. We say that a state \(\mpcstatename'\)
is {\em reachable} if, for some initial state \(\mpcstatename = \mach,\context_0\)
where \(\mach\) is well-formed, \(\mpcstatename \hookrightarrow \MPCT\)
and \(\MPCT\) contains \(\mpcstatename'\).
A {\it reachable segment} \(\MPCT \langle f \rangle\) is a subtrace
such that from some similarly well-formed initial state \(\mpcstatename\),
\(\mpcstatename \hookrightarrow \MPCT \langle f \rangle\).

\section{Stack Safety, Formally}
\label{sec:lse}

Now we can give our definition of stack safety. Each macro-instruction defines a
context step; in turn, the context at the point of a call or return defines a trace
property each for integrity, confidentiality, and well-bracketed control flow.
The context updates involved in calls and returns are given in \ref{fig:callcontexts}.
First, we define a function, {\it update view}, that defines the view of a callee.
Addresses that, offset from stack pointer, are in public regions are marked
\(\mathit{public}\). Those that are being passed on the stack are marked
\(\mathit{passed}\). [When we add pass-by-reference arguments, we wil use
\(V\)]

To update the context, we pair the updated view with the return target of the
call: the

@@

\judgment[]
         {\(j = \mathbf{enter} ~ tr ~ n\)}
         {\(m,c \stepstounder{j}_M c\)}

\judgmentbr[ReturnCCons]
           {\(\)}
           {\(j = \mathbf{return} ~ pubs\)}
           {\(m, V::tars \stepstounder{j}_C ((m[pc] + 4, m[sp]), V')::tars\)}

\judgmentbr[ReturnCNil]
           {\(V' = \lambda \component . \textnormal{if } \component = \PCname \textnormal{ then }
             \unsealed)\)}
           {\(\mathit{decode}^* ~ m ~ pc = \mathit{call} ~ f ~ args ~ sargs ~ pubs\)}
           {\(m, [] \stepstounder{j}_C ((m[pc] + 4, m[sp]),V')::[]\)}

@@

We now state our properties.
The first is {\em stepwise stack integrity}, where ``stepwise'' means that the property is
quantified over each step of execution from any initial state. The criterion for integrity
is simple: sealed elements do not change.

\definition
The \emph{protected set} of a context, \(\prot(\domMap,\_)\), is the
set of elements \(\component\) such that \(\domMap ~ \component = \sealed{\depth}\)
for some depth \(\depth\).

\definition
A system enjoys \textit{\textbf{stepwise stack integrity}} if, for any reachable state
\(\mach,\context\) such that \(\mach,\context \stepstocon \mach',\context'\),
and any element \(\component \in \prot(\context)\),
\(\mach'[\component] = \mach[\component]\).

Next we consider {\em stepwise stack confidentiality}. Recall that our properties are
nested: we will quantify over subtraces that represent whole calls. For each
subtrace, we quantify over {\em variants} of its initial state. Then we require
that the original, {\em primary} traces move in {\em lockstep} with the variant.

\definition The \emph{call set} \(\MPCT^*\) of a machine is
\[\begin{aligned}
\bigcup_{\depth \in \mathbb{Z}} \{(\MPCT, \neg f_\depth) | \MPCT\langle f_\depth \rangle
\text{ is a reachable segment}\} \hspace{1em} \\
\text{where } f_\depth ~ (\mach,(\domMap,\rts)) \triangleq |\rts| \geq \depth \\
\end{aligned}\]

\noindent
This is the set of all possible calls that can be reached from any well-formed initial state.

We term \(\neg f_\depth\) the {\em break condition} \(\brk\), a predicate on
states that holds on \(\MPCT\)'s final state, {\em unless it terminated prematurely due to a failstop}.

@@

\definition The \emph{difference} of two machine states \(\mach\) and \(\mach'\), written \(\Delta(\mach,\mach')\),
is the set of elements \(\component\)
such that \(\mach[\component] \not = \mach'[\component]\).

\definition Two pairs of states \((\mach,\mach')\) and \((\nach,\nach')\)
 {\em change together}, written \((\mach,\mach') \diamond (\nach,\nach')\), if, for all elements \(\component\) in
\(\Delta(\mach,\mach') \cup \Delta(\nach,\nach')\),
\(\mach'[\component] = \nach'[\component]\).

\definition A pair of traces \(\machT\) and \(\nachT\) are {\em in lockstep},
written \(\machT \doteq \nachT\),
if their adjacent pairs of states change together.  We define this as a coinductive relation:

\smallskip
\begin{minipage}[b]{.2\columnwidth}
\judgment{}
         {\(\mpcstatename \doteq \npcstatename\)}
\end{minipage}
\begin{minipage}[b]{.75\columnwidth}
\judgmenttwo{\(\mach, \pi_\mach(\head(\machT)) \diamond \nach, \pi_\mach(\head(\nachT))\)}
            {\(\machT \doteq \nachT\)}
            {\(\notfinished{(\mach,\underscore)}{\machT} \doteq \notfinished{(\nach,\underscore)}{\nachT}\)}
\end{minipage}

\definition
The \emph{secret set} of a context, written \(\secret(\domMap,\rts)\), is the
set of elements \(\component\) such that \(\domMap ~ \component \not = \outside\).

\definition
A system enjoys \textbf{\textit{stepwise stack confidentiality}} if,
for any \((\MPCT,\brk) \in \MPCT^*\) with \(\head(\MPCT) = (\mach,\context\)),
any \(\nach\) such that \(\mach \approx_{\mathit{secret}(\context)} \nach\), and any trace
%\(\NPCT\) such that \(\nach,\pol,\domMap \hookrightarrow \NPCT | \brk\),
\(\NPCT\) such that \((\nach,\context) \hookrightarrow \NPCT | \brk\),
\(\MPCT \doteq \NPCT\) holds.

In short, if we consider a variant of the stack at the start of a call,
the resulting trace remains in lockstep with the original trace
until either both return together or one failstops.
Note that \(\MPCT\) may contain nested calls; these must also respect the secrecy
of the elements in \(\components\), and in addition their subtraces are also in
\(\MPCT^*\), and therefore confidentiality must hold on them separately.

Finally, WBCF is a \ifspace simple \fi property of the behavior of returns.
\definition
A system enjoys \textit{\textbf{well-bracketed control flow}} when, for every reachable state
\(\mpcstatename\) where \(\pi_\context(\mpcstatename) = (\underscore,\rts)\), and
\(\mpcstatename \stepsto \mpcstatename'\) where \(\pi_\context(\mpcstatename') = (\underscore,\rts')\),
if \(|\rts'| < |\rts|\), then \(|\rts'| = |\rts|-1\).

\medskip
\section{Case Study: Micro-Policies}
\label{sec:enforcement}

We next explore how an \ifspace existing \fi enforcement mechanism called {\em
  Depth Isolation}~\cite{DBLP:conf/sp/RoesslerD18} implements
the formal stack safety property described in the previous
section, and why its optimized form, {\em Lazy Tagging and Clearing}, does not.
We validate these claims using property-based random testing in \cref{sec:testing}.

\paragraph*{Conservative Stack Safety Enforcement Mechanism}
%
The enforcement policy of Roessler and DeHon \cite{DBLP:conf/sp/RoesslerD18} relies
on a programmable, tag-based reference monitor that runs alongside the
program,
allowing fine-grained manipulation of metadata tags to encode so-called
\emph{micro-policies}~\cite{pump_oakland2015}.
Every value in
memory and registers (including the $\PCname$)
is enriched with an abstract metadata tag, which can represent
arbitrary information about the value. A micro-policy is then defined as a
set of tags and a
collection of software-defined rules.
On each machine step, the relevant rule is applied to the tags on
the instruction's inputs ($\PCname$, registers, memory) and on the
\ifspace instruction \fi opcode\ifspace itself\fi, producing
one of two outcomes: either the instruction is permitted to execute (and
generates tags for the result of the operation and the new $\PCname$), or
the machine fail-stops with a policy violation.
Existing work \cite{TestingNI:ICFP,pump_oakland2015, DBLP:conf/sp/RoesslerD18}
%\bcp{additional citations: the noninterference paper, Nick and Andre's papers, ...}
has shown that a wide range of
micro-policies can be defined using this scheme.
Efficient execution of these micro-policies relies on hardware acceleration,
such as the PUMP architecture~\cite{pump:asplos2015}.

Tags in the PIPE system are not accessible to application code, being separated
from their associated values at the hardware level. So they are not modeled
as state elements, and therefore our properties make no claims about
their values.

The micro-policy we describe here is a variation
on the {\em Depth Isolation} policy presented by
Roessler and DeHon\cite{DBLP:conf/sp/RoesslerD18}.
The Depth Isolation micro-policy maintains tags of the form $\tagStackDepth{n}$ for
each stack location belonging to the stack frame at activation depth
$\depth$, or the tag $\tagNoDepth$ for stack locations that are currently
unused (but could be used in the future).
%
The $\PCname$ is tagged similarly as $\tagPCDepth{\depth}$.
%
Initially, the entire stack is considered unused (and therefore tagged
$\tagNoDepth$), and the {\PCname} has tag $\tagPCDepth{0}$.
%
During normal execution, the micro-policy rules only permit load and
store operations when the target memory is tagged {\em with the same
 depth} as the current {\PCname} tag, and store operations also when the
target memory is $\tagNoDepth$.

Interesting changes to the tags occur where the program switches from
a caller to a callee or vice-versa:

\begin{itemize}

\item From caller to callee: when the machine executes a call,
  %, when the machine executes an instruction marked as a call in
  % the call map.
  the current
  $\tagPCDepth{\depth}$ tag is incremented to $\tagPCDepth{(\depth + 1)}$, and the function
  entry sequence initializes  all locations of the new function frame with
  tag $\tagStackDepth{(\depth + 1)}$.

\item From callee back to caller: the callee clears its frame, and at
  the actual return instruction,
  the $\PCname$ tag is decremented.
\end{itemize}

To implement this discipline, blessed instruction sequences are designated
to appear at the entry and exit of each function,
which manipulate tags as just described in addition to performing the
usual calling convention tasks of saving/restoring the return address to/from
the stack and adjusting the stack pointer. With the aid of an additional tag on
the $\rsp$ register, these sequences also serve to enforce
well-bracketed control flow.
The micro-policy guarantees atomic execution of these sequences
using a combination of tags on the instructions
and an additional tag on the $\PCname$.

There remains the question of how to ensure that the sequences are
invoked at the right places\ifspace from the perspective of the stack safety
property\fi. For a program to enjoy stack safety
(with respect to a particular set of call annotations), all we require is that
the entry sequence immediately follows any instruction marked as a call
in the call map. This is achieved by giving these instructions (another) special tag,
and it is easy to check statically that this has been done correctly.
If the code fails to initiate an exit
sequence at a point where the stack safety property expects a return,
the micro-policy will incorrectly behave as if execution
is continuing in the callee, but since the callee never has stronger access
rights than the caller, this is harmless.

\medskip
\paragraph*{Lazy Alternative}

The conservative policy described above is rather slow in
some important practical cases. Roessler and DeHon\cite{DBLP:conf/sp/RoesslerD18} observe
that most of the performance overhead incurred
stems from the need to set stack activation tags as a frame is created
and reset them as it is destructed;  this
is particularly costly for frames that contain large uninitialized arrays.
To mitigate those costs, they eliminate the costly frame initialization pass.

The most optimized version of the micro-policy, {\em Lazy Tagging and Clearing},
does not initialize stack frames on entry \emph{or} clear them on exit,
and it permits all writes to the stack, even when the $\PCname$ tag
does not match the memory tag; the \(\PCname\) tag is always propagated
to the written location. Reads from the stack \emph{do} require $\PCname$
tag and memory tag to match. Thus, even if a callee illicitly writes to
a private location in its caller, the caller will eventually detect this
if it ever tries to read from that location.

This \emph{lazy policy} admits a more efficient implementation, but
deliberately allows temporary violations of stack integrity.
The natural question then is: how do we characterize the
protections provided by these policies once stepwise stack safety is broken?
How does a harmful violation that must be caught later differ from a
harmless one?

>>>>

We begin with a RISC-V-like machine enhanced with PIPE, a tag-based reference
monitory. We extend this machine with a {\it security overlay semantics}:
a semantics in which calls, returns, and other operations of interest are condensed
into atomic {\it macro-instructions}, which update an additional security context.
As opposed to a fully-abstract overlay semantics, in which the semantics is safe
by construction and compilation to an enforcement mechanism must be proved fully abstract,
our machine's semantics are not safe unless the tag policy is correctly
instantiated, but the security context informs trace properties that
must hold under a policy for that policy to be considered valid.

@@

Collectively addresses and registers are {\em state elements} \(\component\)
in the set \(\COMPONENTS ::= \WORDS + \REGS\).
%
A {\em machine state} \(\mach \in \MACHS ::= \COMPONENTS \rightarrow \WORDS\)
is a mapping from state elements to values. For convenience, we abbreviate
multiple updates \(m[\component_0 \mapsto \word_0][\component_1 \mapsto \word_1]\dots\)
as \(m \llbracket A \mapsto B | C \rrbracket\), where \(A\) and \(B\)
are fomulae with free variables bound to sets in \(C\).

The machine is equipped with additional state representing the tag-based hardware.
We represent this as a policy state, \(\pol \in \POLS ::= \COMPONENTS \rightarrow \mathcal{T}\),
where \(\mathcal{T}\) is the set of tags, discussed [in the section on security.]
We represent a {\it policy} via the partial function
\(\cdot, \cdot \rightharpoonup_P \cdot \in \MACHS \rightarrow \POLS \rightharpoonup \POLS\),
where the function being undefined represents a {\it failstop}.

@@

Each of these types of state is combined in to a triple called an
MPC-state, \(\mpcstatename \in \MPCS ::= \MACHS \times \POLS \times \CONTEXTS\).
Execution is modeled by a partial step function between MPC-states,
written \(\mach, \pol, \context \harpoonunder{} \mach', \pol', \context'\),
optionally annotated with a macro-instruction \(\harpoonunder{j}\).
If \(\mach[\PCname]\) decodes to an instruction, then execution follows
the RISC-V semantics and the policy, without touching the security context.

\judgmentthree[]
              {\(\mathit{decode} ~ \mach ~ \mach[\PCname] \in I\)}
              {\(\mach \stepstounderfull \mach'\)}
              {\(\mach,\pol \rightharpoonup_P \pol'\)}
              {\(\mach,\pol,\context \harpoonunder{} \mach',\pol',\context\)}

If, for some \(j\), \(\mach[\PCname + i]\) decodes to \(j_i\) for all \(i\) between 0 and
\(\mathit{size}(j)\), written \(\mathit{decode*} = j\), then the step is via a special
multi-step. We present these steps as separate step relations:
\(\cdot \harpoonunder{\cdot}_M \cdot \in \MACHS \times J \times \MACHS\) to step
the machine state,
\(\cdot, \cdot \harpoonunder{\cdot}_P \cdot \in \MACHS \times \POLS \times J \times \POLS\)
for the policy state, and
 \(\cdot, \cdot \stepstounder{j}_C \cdot \in \MACHS \times \CONTEXTS \times J \times \CONTEXTS\)
for the security context. We then combine them:

\judgmenttwobr[]
               {\(\mathit{decode*} = j\)}
               {\(\mach \harpoonunder{j}_M \mach'\)}
               {\(\mach,\pol \rightharpoonup_P \pol'\)}
               {\(\mach',\context \stepstounder{j}_C \context'\)}
               {\(\mach,\pol,\context \harpoonunder{j} \mach',\pol',\context'\)}

In this section we will define \(\harpoonunder{j}_M\) for our machine; \(\harpoonunder{j}_P\)
for the relevant policies will be discussed in [enforcement section], and \(\stepstounder{j}_C\)
in Section \ref{sec:properties}.

@@

Then, on enter, the callee stores the stack pointer and callee-save registers,
and allocates new stack space.

\vspace{\abovedisplayskip}
\judgmentbrbrbr[EnterM]
               {\(j = \mathit{decode}^* ~ m ~ m[pc] = \mathit{enter} ~ tr\)}
               {\(m' = m[m[sp] \mapsto m[rra]]\)}
               {\(m'' = m\llbracket m[m[sp]+i+1 \mapsto m[rs_i] | 0 \leq i < 8 \rrbracket\)}
               {\(m''' = m''[sp \mapsto m[sp]+n+9]]\)}
               {\(m \harpoonunder{j}_M m'''[pc \mapsto m[pc]+4*\mathit{size}(j)]\)}
\vspace{\belowdisplayskip}

Finally, on return, the program counter and register contents are restored from the stack,
and the callee's frame is deallocated.

\vspace{\abovedisplayskip}
\judgmentbrbr[ReturnM]
             {\(j = \mathit{decode}^* ~ m ~ m[pc] = \mathit{return} ~ n\)}
             {\(m' = m[sp \mapsto m[sp]-n-9]\)}
             {\(m'' = m\llbracket m[rs_i] \mapsto m[m[sp]+i+1 | 0 \leq i < 8 \rrbracket\)}
             {\(m \harpoonunder{j}_M m''[pc \mapsto m''[sp]]\)}
\vspace{\belowdisplayskip}

@@

Every macro-instruction is associated with a sequence of normal RISC-V instructions
of the appropriate length, so compiling from the macro-instruction machine to normal
RISC-V is as simple as replacing \(j_i\) with its corresponding instruction, tagged
as a ``blessed sequence'' -- a standard tag-based enforcement mechanism that guarantees
that instructions may {\it only} step in the appropriate order.

\paragraph*{Machine and Observation Traces}
\label{sec:traces}

A {\em machine trace} is a nonempty, finite or infinite sequence
of states, ranged over by \(\MPCT\) and \(\NPCT\).
We use ``\(\notfinished{}{}\)'' to represent ``cons'' for traces, reserving ``::''
for list-cons. An {\em event trace} \(\obsT\) is a similar sequence of
events \(\obs \in \OBSS\), which are defined as follows:

\[\begin{split}
\obs \in \OBSS ::= & ~ \mathbf{callE} ~ a_{dst} ~ \overline{w_{args}} \\
| & ~ \tau \\
\end{split}\]

%This system of observations resembles that used by CompCert [cite], in that
%the observable behaviors of the system are calls and returns (in our case, not
%distinguishing between internal and external, and including exceptions), as well
%as a special output akin to storing to volatile memory. We add the silent observation
%\(\tau\).

We convert macro-instructions into events via the \(\mathit{observe}\)
function, observing the values of relevant registers, defined in Figure \ref{fig:observe}.

\begin{figure*}
\[\mathit{observe} ~ j ~ m =
\begin{cases}
  \mathbf{callE} ~ m[r] ~ m \llbracket k | k \in tr.args \lor m[rsp] - k \in tr.sargs \rrbracket
  & \textnormal{when } ~ j = \mathbf{call} ~ r ~ tr ~ pubs \\
  \tau & \textnormal{otherwise} \\
\end{cases}\]
\label{fig:observe}
\end{figure*}

The ``trace-of'' operator, written \(\mpcstatename \hookrightarrow \MPCT \),
coinductively relates an initial state with the trace of states
produced by repeated application of \(\rightharpoonup\):

\judgmenttwo{\(\mpcstatename \rightharpoonup \mpcstatename'\)}
            {\(\mpcstatename' \hookrightarrow \MPCT \)}
            {\(\mpcstatename \hookrightarrow \notfinished{\mpcstatename'}{\MPCT} \)}%
\judgment%[Default]
         {\(\not\exists \mpcstatename'. \mpcstatename \stepstocon \mpcstatename'\)}
         {\(\mpcstatename \hookrightarrow \mpcstatename\)}

@@

This operator, written \(\obsof\), takes a machine trace and
gives the trace of events of each step in
the trace. It uses a coinductive helper \(\mathit{obsfor}\) that relates an
initial event, a machine-trace, and an event trace; the helper is
then invoked with an initial \(\tau\).

\judgmenttwobrlong{\(\mpcstatename \harpoonunder{j} \mpcstatename'\)}
                  {\(\mpcstatename' \hookrightarrow \MPCT\)}
                  {\(\mathit{obsfor} ~ (\mathit{observe} ~ \pi_m(\mpcstatename) ~ j) ~ \MPCT = \obsT\)}
                  {\(\mathit{obsfor} ~ \obs ~ \notfinished{\mpcstatename}{\MPCT} = \notfinished{\obs}{\obsT}\)}

We say that two event traces $\obsT_1$ and $\obsT_2$ are {\em similar}
from the perspective of an external observer, written \(\obsT_1 \eqsim
\obsT_2\), if the sequence of non-silent events is the same. That is, we
compare up to deletion of \(\tau\) events, coinductively:

Note that an infinite silent trace is a
prefix of (and similar to) any other trace. While this might seem
surprising at first, it makes sense in a timing-insensitive context:
an external observer looking at two machine runs cannot (computably)
distinguish between a machine that steps forever and a machine that
steps for a long time before producing some output.

@@

We often want to quantify over all traces from any initial state so that our
properties apply to the system as a whole. We say that a state \(\mpcstatename'\)
is {\em reachable} if, for some initial state \(\mpcstatename = \mach,\pol,[]\)
where \(\mach\) is well-formed, \(\mpcstatename \hookrightarrow \MPCT\)
and \(\MPCT\) contains \(\mpcstatename'\).

\section{Stack Safety, Formally}
\label{sec:properties}

Now we can give our definition of stack safety. Each macro-instruction defines a
context step; in turn, the context at the point of a call or return defines a trace
property each for integrity, confidentiality, and well-bracketed control flow.
The context updates involved in calls and returns are given in Figure \ref{fig:callcontexts}.
First, we define a function, {\it update view}, that defines the view of a callee.
Addresses that, offset from stack pointer, are in public regions are marked
\(\mathit{public}\). Those that are being passed on the stack are marked
\(\mathit{passed}\). [When I add pass-by-reference arguments, we will use
the recursive \(V\)]

To update the context, we pair the updated view with the return target of the
call: the return target tells us whether a return is to the proper location
which the stack pointer restored, and the view tells us how to treat the
security of the callee's state.

@@

\judgment[EnterC]
         {\(j = \mathbf{enter} ~ tr ~ n\)}
         {\(m,c \stepstounder{j}_M c\)}

\judgmentthree[ReturnCValid]
              {\(m[pc] = a_{pc}\)}
              {\(m[rsp] = a_{sp}\)}
              {\(j = \mathbf{return} ~ pubs\)}
              {\(m, (\mathit{return} ~ a_{pc} ~ a_{sp}, V)::\context \stepstounder{j}_C \context\)}

\judgmenttwobrlong[ReturnCInvalid]
                  {\(m[pc] \not = a_{pc} \lor m[rsp] \not = a_{sp}\)}
                  {\(j = \mathbf{return} ~ pubs\)}
                  {\(V' = V \llbracket r \mapsto \unsealed | r \in \{pc, rsp\} \rrbracket\)}
                  {\(m, (\mathit{return} ~ a_{pc} ~ a_{sp}, V)::\context \stepstounder{j}_C (\mathit{return} ~ a_{pc} ~ a_{sp}, V')::\context\)}

\judgmenttwo[ReturnCNil]
            {\(V' = \lambda \component . \unsealed\)}
            {\(j = \mathbf{return} ~ pubs\)}
            {\(m, [] \stepstounder{j}_C []\)}

@@

\paragraph*{Eventual Integrity}

The intuition behind eventual integrity is that, when control is passed from the
caller to the callee, the caller doesn't necessarily need its data to remain intact --
as long as they are restored by the time control returns to it. We define this as a more
general notion of {\it eventual integrity.}

Let \(R \subseteq \CONTEXTS\) be a predicate on contexts, representing the condition at
which integrity must be restored, and \(I\) be a predicate on states defining what integrity
means. Then we find the first state on which \(R\) holds on the context and assert that
\(I\) also holds on the machine state, with the \(\mathit{when}\) relation.

\judgmenttwo{\(R ~ \context\)}
            {\(I ~ \mach\)}
            {\(\mathit{when} ~ R ~ I ~ (\mach,\pol,\context)\)}

\judgmentthree{\(\neg R ~ \context\)}
              {\((\mach,\pol,\context) \rightharpoonup \mpcstatename\)}
              {\(\mathit{when} ~ R ~ I ~ \mpcstatename\)}
              {\(\mathit{when} ~ R ~ I ~ \mpcstatename\)}

\definition Let \(\components\) be a set of elements and \((\mach,\pol,\context)\)
be the first state of a new call. We define \(R\) to hold on those states
\((\mach',\pol',\context')\) where \(|\context'| < |\context|\). And we define
\(I\) to hold when \(\mach[\component] = \mach'[\component]\) for all elements
\(\component \in \COMPONENTS\). Then \((\mach,\pol,\context)\) enjoys eventual
integrity of \(\components\) if \(\mathit{when} ~ R ~ I ~ (\mach,\pol,\context)\).

Howver, eventual integrity is not sufficient to express the integrity protected by
Lazy Tagging, because it is possible that values in memory are overwritten -- but
are also guaranteed never to be accessible. We need to define a slightly weaker property.

\paragraph*{Variants and Transient Elements}

One or more elements of a given state are {\it transient} if their values
in that state have no observable influence on future execution. We define this
formally as a sort of non-interference: for a state \(\mpcstatename\) and
a set of elements \(\components\), \(\mathit{transient} ~ \mpcstatename ~ \components\)
holds if we can replace the contents of \(\components\) with arbitrary values
and have the resulting state produce an equivalent trace. The state created
by modifying \(\components\) in this way is termed a variant.

@@

\definition An elements set \(\components\) is transient in state \((\mach,\pol,\context)\),
written \(\mathit{transient} ~ (\mach,\pol,\context) ~ \components\), if for all
\(\nach\) such that \(\mach \approx_{\components'} \nach\), if 
\((\mach,\pol,\context) \hookrightarrow \MPCT\) and
\((\nach,\pol,\context) \hookrightarrow \NPCT\), then
\(\obsof(\MPCT) \eqsim \obsof(\NPCT)\).

\definition Let \(\Delta(\mach,\mach')\) be the set of elements \(\component\)
such that \(\mach[\component] \not = \mach'[\component]\).

\definition Let \(\components\) be a set of elements and \((\mach,\pol,\context)\)
be the first state of a new call. We define \(R\) to hold on those states
\((\mach',\pol',\context')\) where \(|\context'| < |\context|\). And we define
\(I\) to hold on \(\mach'\) when \(\Delta(\mach,\mach') \cap \components\) is transient
with respect to \(\mach'\). Then \((\mach,\pol,\context)\) enjoys
{\it eventual intransient integrity} of \(\components\) if
\(\mathit{when} ~ R ~ I ~ (\mach,\pol,\context)\).

\definition A system enjoys {\it observational stack integrity} if, for all
reachable states \(\mpcstatename\) such that
\(\mpcstatename \harpoonunder{\mathbf{Call}\dots} (\mach,\pol,(tar, V)::\context)\),
\((\mach, \pol, (tar, V)::\context)\) enjoys eventual intransient integrity with
respect to the set of components that \(V\) maps to \(\sealed\).

\paragraph*{Temporary Confidentiality}

Confidentiality properties are also modeled as non-interference, but with a twist:
the caller's confidential data should be kept secret from the callee, but after return,
will naturally be accessible to the caller once again. So, we must model confidentiality
in terms of events inside the callee. But confidentiality violations might not become
visible during the callee's lifetime -- the callee might read a secret and stash it somewhere
that it will be accessed later. So we must also consider the machine state at the end
of the call. First we'll define a useful general property that implements this structure,
similar to \(\mathit{when}\), called \(\mathit{during\_both}\).

\(\mathit{during\_both}\) relates two initial states in their traces up until they
both reach an end condition, again defines by a predicate \(R\). Instead of \(I\)
we have an event-trace relation \(E\) that must relate the event traces up through
\(R\), and a final relation \(F\) that must relate the final states.

\definition Let \(\mpcstatename\) and \(\npcstatename\) be initial states
and \(R\) be a predicate on states such that
\(\mpcstatename \hookrightarrow \MPCT | R\) and
\(\npcstatename \hookrightarrow \NPCT | R\).
Let \(E\) be a relation on event traces, and \(F\) be a relation on machine states.
Then \(\mathit{during\_both} ~ \mpcstatename ~ \npcstatename ~ R ~ E ~ F\) holds if both:
\begin{itemize}
\item \(E ~ (\obsof \MPCT) ~ (\obsof \NPCT)\) holds
\item And, if \(\last ~ \MPCT = (\mach,\_,\_)\) and \(\last ~ \NPCT = (\nach,\_,\_)\),
  then \(F ~ \mach ~ \nach\) holds.
\end{itemize}

We could define a natural, strict form of confidentiality that requires that the final
states be identical. However, as with integrity, this is stricter than we need -- there
may be some conditions in which a secret is copied, but never used again. So, we once
again wish for our final condition to make the values {\it transient}. Which values?
